---
title: "HW #1"
subtitle: "Interpreting `{ggplot2}` code"
description: "Assigned Wed 01/07/2026 | Due Wed 01/14/2026"
author: "Your Name"
toc: true
editor_options: 
  chunk_output_type: console
---

::: {.callout-tip}
## Some notes before you get started
- **Be sure to install any packages** in the Setup chunk that you don't already have.
- **Leave the code chunk options, `eval: false` and `echo: true`, set as they are.** The final infographic has been intentionally optimized (e.g., text size, spacing) for saving and viewing as a PNG file, not for display in the Plots pane or within a rendered Quarto document. As a result, the text in each individual ggplot may appear too large when viewed in the Plots pane, but will be correctly sized in the exported PNG. We’ll talk more about the nuances of saving ggplots (and why these differences occur) in a later lab section.
- Some answers may become clearer once you’ve looked ahead at the code further down in the script. **Consider revisiting questions as you go.**
:::

## I. Setup

```{r}
#| eval: false
#| echo: true

library(colorspace)
library(geofacet) 
library(ggtext) 
library(glue) 
library(grid)
library(magick)
library(patchwork) 
library(scales) 
library(showtext) 
library(tidyverse) 

ufo_sightings <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/ufo_sightings.csv')
places <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2023/2023-06-20/places.csv')

alien <- c("#101319", "#28ee85")
bg <- alien[1]
accent <- alien[2]

ufo_image <- magick::image_read(path = here::here("images", "ufo.png")) 

sysfonts::font_add_google(name = "Orbitron", family = "orb")
sysfonts::font_add_google(name = "Barlow", family = "bar")

sysfonts::font_add(family = "fa-brands", regular = here::here("fonts", "Font Awesome 6 Brands-Regular-400.otf"))
sysfonts::font_add(family = "fa-solid", regular = here::here("fonts", "Font Awesome 6 Free-Solid-900.otf"))

showtext::showtext_auto(enable = TRUE)
```

1. **What is the author defining in lines 15-17? Where else in the code do these defined variables show up? What advantage(s) is there to defining these values here, as variables, rather than defining the values directly throughout the script?**

    - Answer here

2. **In your own words, explain what the function, `font_add_google()`, does. What's the difference between the two arguments, `name` and `family`?**

    - Answer here

## II. Data wrangling

### i. Create `df_pop`

```{r}
#| eval: false
#| echo: true

df_pop <- places |>
  filter(country_code == "US") |>
  mutate(state = str_replace(string = state,
                             pattern = "Fl",
                             replacement = "FL")) |> 
  group_by(state) |>
  summarise(pop = sum(population)) |>
  ungroup()
```

3. **Describe what this data frame contains.**

    - Answer here

### ii. Create `df_us`

```{r}
#| eval: false
#| echo: true

df_us <- ufo_sightings |>
  filter(country_code == "US") |>
  mutate(state = str_replace(string = state,
                             pattern = "Fl",
                             replacement = "FL")) |> 
  count(state) |>
  left_join(df_pop, by = "state") |>
  rename(num_obs = n) |> 
  mutate(
    num_obs_per10k = num_obs / pop * 10000,
    opacity_val = num_obs_per10k / max(num_obs_per10k)
    )
```

4. **Describe what this data frame contains.**

    - Answer here

5. **What does `opacity_val` represent, and why is it calculated?**

    - Answer here

### iii. Create `df_shape`

```{r}
#| eval: false
#| echo: true

df_shape <- ufo_sightings |>
  filter(!shape %in% c("unknown", "other")) |>
  count(shape) |>
  rename(total_sightings = n) |> 
  arrange(desc(total_sightings)) |>
  slice_head(n = 10) |>
  mutate(
    shape = fct_reorder(.f = shape, 
                        .x = total_sightings), 
    opacity_val = scales::rescale(x = total_sightings, 
                                  to = c(0.3, 1))
    )
```

6. **Describe what this data frame contains.**

    - Answer here

7. **What does `fct_reorder` do when it is applied to the `shape` variable? What would have happened if this step was not performed?**

    - Answer here

8. **What is the purpose of rescaling `opacity_val`? And why rescale from 0.3 to 1?**

    - Answer here

### iv. Create `df_day_hour`

```{r}
#| eval: false
#| echo: true

df_day_hour <- ufo_sightings |>
  mutate(
    day = wday(reported_date_time), 
    hour = hour(reported_date_time), 
    wday = wday(reported_date_time, label = TRUE) 
  ) |>
  count(day, wday, hour) |>
  rename(total_daily_obs = n) |> 
  mutate(
    opacity_val = total_daily_obs / max(total_daily_obs),
    hour_lab = case_when(
      hour == 0 ~ "12am",
      hour <= 12 ~ paste0(hour, "am"),
      hour == 12 ~ "12pm",
      TRUE ~ paste0(hour - 12, "pm")) 
    )
```

9. **Describe what this data frame contains.**

    - Answer here

10. **What is the purpose of the last line inside the `case_when()` statement (`TRUE ~ paste0(hour - 12, "pm")`)?**

    - Answer here

## III. Prepare text elements

```{r}
#| eval: false
#| echo: true

quotes <- paste0('"...', str_to_sentence(ufo_sightings$summary[c(47816, 6795, 93833)]), '..."')

original <- glue("Original visualization by Dan Oehm:")
dan_github <- glue("<span style='font-family:fa-brands;'>&#xf09b;</span> doehm/tidytues")
new <- glue("Updated version by Sam Shanny-Csik for EDS 240:")
link <- glue("<span style='font-family:fa-solid;'>&#xf0c1;</span> eds-240-data-viz.github.io")
space <- glue("<span style='color:{bg};'>. .</span>")
caption <- glue("{original}{space}{dan_github}
                <br><br>
                {new}{space}{link}")
```

11. **In your own words, what is the difference between `paste0()` and `glue()`? Why did the author use `paste0` to construct `quotes` and `glue` to construct the other text elements?**

    - Answer here

## IV. Build plots

### i. Build `plot_shape`

```{r}
#| eval: false
#| echo: true

plot_shape <- ggplot(data = df_shape) +
  geom_col(aes(x = total_sightings, y = shape, alpha = opacity_val), 
           fill = accent) +
  geom_text(aes(x = 200, y = shape, label = str_to_title(shape)), 
            family = "orb", 
            fontface = "bold",
            color = bg, 
            size = 14, 
            hjust = 0,
            nudge_y = 0.2) +
  geom_text(aes(x = total_sightings-200, y = shape, label = scales::comma(total_sightings)),
            family = "orb",
            fontface = "bold",
            color = bg,
            size = 10,
            hjust = 1,
            nudge_y = -0.2) +
  scale_x_continuous(expand = c(0, NA)) +
  labs(subtitle = "10 most commonly reported shapes") +
  theme_void() +
  theme(
    plot.subtitle = element_text(family = "bar", 
                                 size = 40, 
                                 color = accent,
                                 hjust = 0,  
                                 margin = margin(b = 10)),
    legend.position = "none" 
  )
```

12. **Explain the values provided to the `x` aesthetic for both text geoms (`shape` & `total_sightings`).**

    - Answer here

### ii. Build `plot_us` 

**HINT:** Consider temporarily commenting out / rearranging the `geom_*()` layers to better understand how this plot is constructed

```{r}
#| eval: false
#| echo: true

plot_us <-  ggplot(df_us) +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 1, alpha = opacity_val), 
            fill = accent) +
  geom_text(aes(x = 0.5, y = 0.7, label = state), 
            family = "orb", 
            fontface = "bold",
            size = 9, 
            color = bg) +
  geom_text(aes(x = 0.5, y = 0.3, label = round(num_obs_per10k, 1)), 
            family = "orb", 
            fontface = "bold",
            size = 8, 
            color = bg) +  
  geofacet::facet_geo(~state) +
  coord_fixed(ratio = 1) +
  labs(subtitle = "Sightings per 10k population") +
  theme_void() +
  theme(
    strip.text = element_blank(),
    plot.subtitle = element_text(family = "bar", 
                                 size = 40, 
                                 color = accent,
                                 hjust = 1,  
                                 margin = margin(b = 10)),
    legend.position = "none" 
  )
```

13. **Consider the order of `geom_*()` layers in the the above plot (`plot_us`). Why did the author order the layers in this way?**

    - Answer here

### iii. Build `plot_day`

```{r}
#| eval: false
#| echo: true

plot_day <- ggplot(data = df_day_hour) +
  geom_tile(aes(x = hour, y = day, alpha = opacity_val), 
            fill = accent, 
            height = 0.9, 
            width = 0.9) +
  geom_text(aes(x = hour, y = 9, label = hour_lab), 
            family = "orb",
            color = accent, 
            size = 10) +
  geom_text(aes(x = 0, y = day, label = str_sub(string = wday, start = 1, end = 1)), 
            family = "orb", 
            fontface = "bold",
            color = bg, 
            size = 8) + 
  ylim(-5, 9) +
  xlim(NA, 23.55) +
  coord_polar() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = bg, color = bg),
    legend.position = "none"
  )
```

14. **This plot includes one-letter labels for each day of the week. How is ths accomplished when week days are written using their three-letter abbreviations (e.g. `Mon`, `Tue`) in the `df_day_hour` data frame?**

    - Answer here

15. **What role do the `ylim()` and `xlim()` functions play in shaping a ggplot, and how do they change the visual layout of this particular plot? To better understand their effect, try rerunning the code with each of these lines commented out and observe how the plot’s spacing and composition change.**

    - Answer here

### iv. Build `quote*`s

A comment from Dan Oehm's original code: "A bit clunky but the path of least resistance."

```{r}
#| eval: false
#| echo: true

quote1 <- ggplot() +
  annotate(geom ="text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[1], width = 40),
           family = "bar", 
           fontface = "italic", 
           color = accent, 
           size = 16, 
           hjust = 0, 
           lineheight = 0.4) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  coord_cartesian(clip = "off")

quote2 <- ggplot() +
  annotate(geom = "text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[2], width = 25),
           family = "bar", 
           fontface = "italic",
           color = accent, 
           size = 16, 
           hjust = 0,  
           lineheight = 0.4) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  coord_cartesian(clip = "off")

quote3 <- ggplot() +
  annotate(geom = "text", 
           x = 0, 
           y = 1, 
           label = str_wrap(string = quotes[3], width = 25),
           family = "bar", 
           fontface = "italic",
           color = accent, 
           size = 16, 
           hjust = 0,  
           lineheight = 0.4) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  coord_cartesian(clip = "off")
```

16. **Why do you think the author chose to convert these text elements (and also in `plot_ufo`, below!) into ggplot objects (you may consider returning to this question after you've worked your way through all of the code)?**

    - Answer here

### v. Build `plot_ufo`

**Note:** Grob stands for **gr**aphical **ob**ject. Each visual element rendered in a a ggplot (e.g. lines, points, axes, entire panels, even images) is represented as a grob. Grobs can be manipulated individually to fully customize plots. 

```{r}
#| eval: false
#| echo: true
plot_ufo <- ggplot() +
  annotation_custom(grid::rasterGrob(ufo_image)) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = bg, color = bg) 
  )
```

### vi. Build `plot_base` 

```{r}
#| eval: false
#| echo: true

plot_base <- ggplot() +
  labs(
    title = "UFO Sightings",
    subtitle = "Summary of over 88k reported sightings across the US",
    caption = caption
    ) +
  theme_void() +
  theme(
    text = element_text(family = "orb", 
                        size = 48, 
                        lineheight = 0.3, 
                        color = accent),
    plot.background = element_rect(fill = bg, 
                                   color = bg),
    plot.title = element_text(size = 128, 
                              face = "bold", 
                              hjust = 0.5, 
                              margin = margin(b = 10)),
    plot.subtitle = element_text(family = "bar", 
                                 hjust = 0.5, 
                                 margin = margin(b = 20)),
    plot.caption = ggtext::element_markdown(family = "bar",
                                            face = "italic",
                                            color = colorspace::darken(accent, 0.25),
                                            hjust = 0.5,
                                            margin = margin(t = 20)),
    plot.margin = margin(b = 20, t = 50, r = 50, l = 50)
  )
```

17. **Why does the author render `plot.caption` using `ggtext::element_markdown()`, rather than `element_text()` (like he does for rendering `plot.title` and `text`)?**

    - Answer here

## V. Assemble & save

```{r}
#| eval: false
#| echo: true

plot_final <- plot_base +
  inset_element(plot_shape, left = 0, right = 1, top = 1, bottom = 0.66) +
  inset_element(plot_us, left = 0.42, right = 1, top = 0.74, bottom = 0.33) +
  inset_element(plot_day, left = 0, right = 0.66, top = 0.4, bottom = 0) +
  inset_element(quote1, left = 0.5, right = 1, top = 0.8, bottom = 0.72) +
  inset_element(quote2, left = 0, right = 1, top = 0.52, bottom = 0.4) +
  inset_element(quote3, left = 0.7, right = 1, top = 0.2, bottom = 0) +
  inset_element(plot_ufo, left = 0.25, right = 0.41, top = 0.23, bottom = 0.17) + 
  plot_annotation(
    theme = theme(
      plot.background = element_rect(fill = bg,
                                     color = bg)
    )
  ) 

ggsave(plot = plot_final, 
       filename = here::here("outputs", "ufo_sightings_infographic.png"), 
       height = 16, 
       width = 10)
```

18. **Explain how `plot_final` is assembled. What do you think is the most challenging aspect of arranging all components into a single plot?**

    - Answer here

19. **Can you think of one reason the author may have chosen to separate the construction of `plot_base` and `plot_final`?**

    - Answer here

## Answer some final reflective questions 

20. **During week 2, we discuss [Choosing the right graphic form](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture2.1-choosing-graphic-forms-slides.html#/title-slide). Refer to this lecture when answering the sub-questions, below:**

    a. **What "perceptual tasks" (from Cleveland & McGill's heirarchy) must the viewer perform to extract information from these visualizations?**
        
        - Answer here
    
    b. **What task(s) do you think the author wanted to enable or message(s) he wanted to convey with these visualizations (see lecture 2.1, slide 16 for examples)? Be sure to note at least one task / message for each of the three data viz.**
    
        - Answer here
    
    c. **Name at least one caveat to the "hierarchy of perceptual tasks" that the author employed to achieve a goal(s) you noted in question b?**
    
        - Answer here

21. **Describe two elements of this piece that you find visually-pleasing / easy to understand / intuitive. Why?** 

    - Answer here

22. **Describe two elements of this piece that you feel could be better presented in a different way. Why?** 

    - Answer here
    
23. **Describe two new things that you learned by interpreting / annotating this code. These could be packages, functions, or even code organizational approaches that you hadn't previously known about or considered.**

    - Answer here

24. **How, if at all, did you use AI tools to help you interpret this code? Describe your approach to using these tools for this assignment. In what ways was consulting the documentation more (or less) helpful than using AI?**

    - Answer here